---
title: "Case Study: Eisenia fetida Growth and Folsomia candida Reproduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case Study: Eisenia fetida Growth and Folsomia candida Reproduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = FALSE
)
```

## Overview

This vignette walks through a complete **BayesianDEB** workflow using the
two bundled datasets that represent typical ecotoxicological test
organisms:

1. **Eisenia fetida** (earthworm) — 21 individuals measured weekly over
   12 weeks.  We fit an individual model, a hierarchical model with
   partial pooling, and compute derived quantities.

2. **Folsomia candida** (springtail) — reproduction test with
   5 cadmium concentrations.  We fit a DEBtox model, estimate EC50/NEC,
   and produce a dose-response curve.

Both analyses follow the standard BDEB workflow:

> data → model → fit → diagnose → interpret

## Prerequisites

```{r prerequisites}
library(BayesianDEB)
library(ggplot2)

# Verify that cmdstanr and CmdStan are available
check_cmdstanr()
```


# Part 1: Eisenia fetida Growth

## 1.1 Explore the data

The `eisenia_growth` dataset contains simulated weekly length
measurements for 21 *Eisenia fetida* individuals over 84 days.

```{r eisenia-explore}
data(eisenia_growth)

str(eisenia_growth)
# 273 obs. of 3 variables: id, time, length

# How many individuals? How many time points?
length(unique(eisenia_growth$id))   # 21
length(unique(eisenia_growth$time)) # 13 (days 0, 7, ..., 84)

# Quick visual
ggplot(eisenia_growth, aes(time, length, group = id)) +
  geom_line(alpha = 0.3) +
  geom_point(size = 0.8, alpha = 0.4) +
  theme_bw(base_size = 12) +
  labs(x = "Time (days)", y = "Structural length (cm)",
       title = "Eisenia fetida: 21 individuals, 12 weeks")
```

Key observations:

- All individuals start at approximately 0.1 cm.
- Growth follows the expected DEB sigmoidal pattern.
- There is visible individual variation, especially in growth rate.

## 1.2 Individual model (single organism)

We start with a single individual to validate the model before scaling
up to the hierarchical version.

### Prepare data

```{r ind-data}
# Pick individual #5
df1 <- eisenia_growth[eisenia_growth$id == 5, ]

dat1 <- bdeb_data(growth = df1, f_food = 1.0)
dat1
```

### Specify model with informative priors

We use biological knowledge to set priors.  For *E. fetida* under
standard lab conditions:

- $\{p_{Am}\}$ (assimilation rate): values around 3–8 J/d/cm² are
  typical → `prior_lognormal(mu = 1.5, sigma = 0.5)` centres the
  prior around $e^{1.5} \approx 4.5$.
- $\kappa$: earthworms invest a large fraction in soma →
  `prior_beta(a = 3, b = 2)` favours values around 0.6.
- $\sigma_L$: measurement precision is ~0.02 cm →
  `prior_halfnormal(sigma = 0.05)`.

```{r ind-model}
mod1 <- bdeb_model(dat1, type = "individual",
  priors = list(
    p_Am    = prior_lognormal(mu = 1.5, sigma = 0.5),
    p_M     = prior_lognormal(mu = -1.0, sigma = 0.5),
    kappa   = prior_beta(a = 3, b = 2),
    v       = prior_lognormal(mu = -1.5, sigma = 0.5),
    E_G     = prior_lognormal(mu = 6.0, sigma = 0.5),
    sigma_L = prior_halfnormal(sigma = 0.05)
  ))
mod1
```

### Fit

```{r ind-fit}
fit1 <- bdeb_fit(mod1,
  chains         = 4,
  iter_warmup    = 1000,
  iter_sampling  = 2000,
  adapt_delta    = 0.9,
  seed           = 42
)
fit1
```

### Diagnose convergence

```{r ind-diag}
diag1 <- bdeb_diagnose(fit1)

# Visual checks
plot(fit1, type = "trace")
plot(fit1, type = "trace", pars = c("p_Am", "kappa", "sigma_L"))
```

What to look for:

- **Zero divergences** — if present, increase `adapt_delta`.
- **All $\hat{R} < 1.01$** — chains have mixed.
- **Bulk ESS > 400** — enough effective samples.

### Posterior summary

```{r ind-summary}
bdeb_summary(fit1, prob = 0.90)

# Focus on key DEB parameters
bdeb_summary(fit1,
  pars = c("p_Am", "p_M", "kappa", "v", "E_G", "sigma_L"),
  prob = 0.95)
```

### Posterior correlations

```{r ind-pairs}
plot(fit1, type = "pairs",
     pars = c("p_Am", "p_M", "kappa", "E_G"))
```

Strong correlations between $\{p_{Am}\}$ and $[p_M]$ are expected: both
control the ultimate size.

### Posterior predictive check

```{r ind-ppc}
ppc1 <- bdeb_ppc(fit1, type = "growth")
ppc1
plot(ppc1, n_draws = 200)
```

If the observed data (red points) fall within the cloud of replicated
trajectories (grey lines), the model fits well.

### Trajectory plot

```{r ind-traj}
plot(fit1, type = "trajectory", n_draws = 200)
```

### Derived quantities

```{r ind-derived}
der1 <- bdeb_derived(fit1,
  quantities = c("L_inf", "k_M", "growth_rate"),
  f = 1.0)

posterior::summarise_draws(der1,
  "mean", "sd",
  "q5"  = ~quantile(.x, 0.05),
  "q95" = ~quantile(.x, 0.95))
```

- **L_inf**: ultimate structural length under ad libitum feeding.
- **k_M**: somatic maintenance rate constant ($[p_M]/[E_G]$).
- **growth_rate**: von Bertalanffy growth rate.

### Predictions under reduced food

```{r ind-derived-food}
# What if food drops to 70%?
der1_f07 <- bdeb_derived(fit1, quantities = "L_inf", f = 0.7)

posterior::summarise_draws(der1_f07, "mean", "sd",
  "q5" = ~quantile(.x, 0.05), "q95" = ~quantile(.x, 0.95))
```


## 1.3 Hierarchical model (all 21 individuals)

The hierarchical model estimates population-level distributions for
$\{p_{Am}\}$ while sharing $[p_M]$, $\kappa$, $v$, $[E_G]$ across
individuals.  This is the preferred approach when you have multiple
organisms.

### Prepare data

```{r hier-data}
dat_all <- bdeb_data(growth = eisenia_growth, f_food = 1.0)
dat_all
# 21 individuals, 273 observations
```

### Specify model

```{r hier-model}
mod_h <- bdeb_model(dat_all, type = "hierarchical",
  priors = list(
    mu_log_p_Am    = prior_normal(mu = 1.5, sigma = 0.5),
    sigma_log_p_Am = prior_exponential(rate = 2),
    p_M            = prior_lognormal(mu = -1.0, sigma = 0.5),
    kappa          = prior_beta(a = 3, b = 2),
    v              = prior_lognormal(mu = -1.5, sigma = 0.5),
    E_G            = prior_lognormal(mu = 6.0, sigma = 0.5),
    sigma_L        = prior_halfnormal(sigma = 0.05)
  ))
mod_h
```

### Fit

Hierarchical models need higher `adapt_delta` because the
population-level and individual-level parameters can create challenging
geometry for the sampler.

```{r hier-fit}
fit_h <- bdeb_fit(mod_h,
  chains         = 4,
  iter_warmup    = 1000,
  iter_sampling  = 2000,
  adapt_delta    = 0.95,
  max_treedepth  = 12,
  seed           = 123
)
```

### Diagnose

```{r hier-diag}
bdeb_diagnose(fit_h)

# Population-level trace
plot(fit_h, type = "trace",
  pars = c("mu_log_p_Am", "sigma_log_p_Am"))

# Posterior densities
plot(fit_h, type = "posterior",
  pars = c("mu_log_p_Am", "sigma_log_p_Am", "p_M", "kappa"))
```

### Population summary

```{r hier-pop}
# Population parameters
bdeb_summary(fit_h,
  pars = c("mu_log_p_Am", "sigma_log_p_Am",
           "p_M", "kappa", "v", "E_G", "sigma_L"),
  prob = 0.95)
```

### Individual assimilation rates

The hierarchical model produces a **shrinkage** effect: individuals
with sparse or noisy data are pulled toward the population mean.

```{r hier-ind}
# All 21 individual p_Am values
bdeb_summary(fit_h,
  pars = paste0("p_Am_ind[", 1:21, "]"),
  prob = 0.90)

# Compare with the population mean
pop_mean <- exp(
  posterior::summarise_draws(
    posterior::as_draws_df(fit_h$fit$draws()),
    "mean"
  )$mean[1]  # mu_log_p_Am
)
cat("Population mean p_Am:", round(pop_mean, 2), "\n")
```

### Prediction for a new individual

The Stan model generates `p_Am_new` in the `generated quantities`
block — a draw from the population distribution, useful for predicting
a yet-unobserved individual.

```{r hier-new}
bdeb_summary(fit_h, pars = "p_Am_new", prob = 0.95)
```

### Compare individual vs hierarchical

```{r compare}
# Individual fit (single organism):
s1 <- bdeb_summary(fit1,
  pars = c("p_Am", "p_M", "kappa"),
  prob = 0.90)
cat("Individual model (id=5):\n")
print(as.data.frame(s1), digits = 3, row.names = FALSE)

# Hierarchical fit (population + individual):
s_h <- bdeb_summary(fit_h,
  pars = c("p_M", "kappa"),
  prob = 0.90)
cat("\nHierarchical model (shared parameters):\n")
print(as.data.frame(s_h), digits = 3, row.names = FALSE)
```

The hierarchical model typically gives **tighter credible intervals**
for shared parameters because it pools information across 21 individuals.


# Part 2: Folsomia candida Reproduction (DEBtox)

## 2.1 Explore the data

The `folsomia_repro` dataset is a standard 28-day reproduction test
with *Folsomia candida* exposed to cadmium.

```{r folsomia-explore}
data(folsomia_repro)

str(folsomia_repro)
# 30 obs: 5 concentrations x 6 replicates

# Offspring counts by concentration
ggplot(folsomia_repro,
       aes(x = factor(concentration), y = count)) +
  geom_boxplot(fill = "steelblue", alpha = 0.3) +
  geom_jitter(width = 0.15, size = 2, alpha = 0.6) +
  theme_bw(base_size = 12) +
  labs(x = "Cd concentration (mg/kg)",
       y = "Offspring count (28 days)",
       title = "Folsomia candida: 28-day reproduction test")
```

Key observations:

- Control (0 mg/kg): ~110–130 offspring.
- 10 and 50 mg/kg: slight reduction.
- 100 mg/kg: clear reduction to ~60–95.
- 500 mg/kg: near-complete inhibition (0–16 offspring).

## 2.2 Prepare data for DEBtox

The DEBtox model needs both growth data and concentration mapping.
Since `folsomia_repro` only has reproduction data (no individual growth
curves), we combine it with simulated growth from `debtox_growth` as a
demonstration.  In a real study, you would supply your own measured
growth data.

For this case study, we use `debtox_growth` which has 4 concentration
groups with growth data:

```{r debtox-data}
data(debtox_growth)

# Visualise growth under toxicant
ggplot(debtox_growth,
       aes(time, length, colour = factor(concentration),
           group = id)) +
  geom_line(alpha = 0.3) +
  geom_point(size = 0.8, alpha = 0.4) +
  facet_wrap(~concentration, labeller = label_both) +
  theme_bw(base_size = 11) +
  labs(x = "Time (days)", y = "Structural length (cm)",
       colour = "Concentration",
       title = "DEBtox growth data: 4 concentration groups")
```

### Build data object

```{r debtox-prep}
# Concentration mapping: match IDs to concentration groups
conc_levels <- unique(debtox_growth$concentration)
conc_map <- setNames(conc_levels, as.character(conc_levels))

dat_tox <- bdeb_data(
  growth        = debtox_growth,
  concentration = conc_map,
  f_food        = 1.0
)
dat_tox
```

## 2.3 Specify DEBtox model

We use `bdeb_tox()` which is a convenience wrapper for `bdeb_model()`
with `type = "debtox"`.  The current implementation supports stress on
assimilation.

```{r debtox-model}
mod_tox <- bdeb_tox(dat_tox, stress = "assimilation",
  priors = list(
    p_Am    = prior_lognormal(mu = 1.5, sigma = 0.5),
    p_M     = prior_lognormal(mu = -1.0, sigma = 0.5),
    kappa   = prior_beta(a = 3, b = 2),
    v       = prior_lognormal(mu = -1.5, sigma = 0.5),
    E_G     = prior_lognormal(mu = 6.0, sigma = 0.5),
    sigma_L = prior_halfnormal(sigma = 0.05),
    # Toxicological priors
    k_d     = prior_lognormal(mu = -1.0, sigma = 1.0),
    z_w     = prior_lognormal(mu = 2.5, sigma = 1.0),
    b_w     = prior_lognormal(mu = -5.0, sigma = 2.0)
  ))
mod_tox
```

Prior rationale for tox parameters:

- **k_d** (damage recovery rate): `prior_lognormal(mu=-1, sigma=1)` centres
  around $e^{-1} \approx 0.37$ d$^{-1}$, allowing fast or slow recovery.
- **z_w** (NEC): `prior_lognormal(mu=2.5, sigma=1)` centres around
  $e^{2.5} \approx 12$, which is sensible given the data range 0–200.
- **b_w** (effect intensity): `prior_lognormal(mu=-5, sigma=2)` is a
  weakly informative prior on a small positive value.

## 2.4 Fit

DEBtox models solve an ODE per concentration group per iteration, so
they are computationally heavier.  High `adapt_delta` is recommended.

```{r debtox-fit}
fit_tox <- bdeb_fit(mod_tox,
  chains         = 4,
  iter_warmup    = 1000,
  iter_sampling  = 2000,
  adapt_delta    = 0.95,
  max_treedepth  = 12,
  seed           = 77
)
```

## 2.5 Diagnose

```{r debtox-diag}
bdeb_diagnose(fit_tox)

# Trace: DEB parameters
plot(fit_tox, type = "trace",
  pars = c("p_Am", "p_M", "kappa"))

# Trace: Tox parameters
plot(fit_tox, type = "trace",
  pars = c("k_d", "z_w", "b_w"))

# Posterior densities
plot(fit_tox, type = "posterior",
  pars = c("k_d", "z_w", "b_w"))

# Pairs: check for pathological correlations
plot(fit_tox, type = "pairs",
  pars = c("z_w", "b_w", "k_d"))
```

## 2.6 Tox parameter summary

```{r debtox-summary}
bdeb_summary(fit_tox,
  pars = c("p_Am", "p_M", "kappa", "v", "E_G",
           "k_d", "z_w", "b_w", "sigma_L"),
  prob = 0.95)
```

## 2.7 EC50 and NEC

The key ecotoxicological endpoints.  The EC50 is computed analytically
in Stan as $\text{EC}_{50} = z_w + 0.5/b_w$.

```{r debtox-ec50}
ec <- bdeb_ec50(fit_tox, prob = 0.95)

# Summary table
print(ec$summary, digits = 3)

# Posterior distribution of EC50
hist(ec$draws, breaks = 50, col = "steelblue", border = "white",
     main = "Posterior distribution of EC50",
     xlab = "Concentration")
abline(v = ec$summary$median[1], col = "red", lwd = 2, lty = 2)
legend("topright", "Posterior median", col = "red", lty = 2, lwd = 2)
```

**Interpretation:**

- The NEC (No Effect Concentration) represents the threshold below
  which no toxic effect is expected.
- The EC50 is the concentration causing 50% reduction in the affected
  process (here, assimilation).
- The full posterior gives you a probability distribution, not just a
  point estimate — crucial for regulatory risk assessment.

## 2.8 Dose-response curve

```{r debtox-dr}
plot_dose_response(fit_tox, n_draws = 200)
```

This plot shows the stress function $\max(1 - b_w\max(C-z_w, 0), 0)$
evaluated at each posterior draw.  The observed data points are
normalised to the control group.

## 2.9 Prior sensitivity analysis

Always check whether the data are informative by comparing posteriors
under different priors:

```{r debtox-sens}
# Tighter prior on z_w
mod_tox2 <- bdeb_tox(dat_tox, stress = "assimilation",
  priors = list(
    z_w = prior_lognormal(mu = 3.0, sigma = 0.3),  # tighter
    b_w = prior_lognormal(mu = -5.0, sigma = 2.0)
  ))
fit_tox2 <- bdeb_fit(mod_tox2, chains = 4, adapt_delta = 0.95,
                     seed = 78)

# Compare
cat("Original z_w prior: LogNormal(2.5, 1.0)\n")
bdeb_summary(fit_tox,  pars = c("z_w", "b_w"), prob = 0.95)

cat("\nTighter z_w prior: LogNormal(3.0, 0.3)\n")
bdeb_summary(fit_tox2, pars = c("z_w", "b_w"), prob = 0.95)
```

If the posteriors agree despite different priors, the data are driving
the inference (good).  If they differ substantially, the parameter is
prior-dominated and you should report this.


# Part 3: Utility Functions

## 3.1 Arrhenius temperature correction

If your experiments were run at a temperature different from the
reference, correct rate parameters before interpreting them:

```{r arrhenius}
# Experiment at 22°C, reference is 20°C
cT <- arrhenius(T = 273.15 + 22, T_ref = 273.15 + 20, T_A = 8000)
cat("Temperature correction factor:", round(cT, 3), "\n")

# To get rate at reference temperature:
# p_Am_ref = p_Am_observed / cT
```

## 3.2 Energy flux calculator

Explore the energetics at a specific state:

```{r fluxes}
fl <- deb_fluxes(
  E = 10, V = 0.5, f = 1.0,
  p_Am = 5, p_M = 0.5, kappa = 0.75,
  v = 0.2, E_G = 400
)

cat("Assimilation:  ", round(fl$p_A, 3), "J/d\n")
cat("Mobilisation:  ", round(fl$p_C, 3), "J/d\n")
cat("Maintenance:   ", round(fl$p_M, 3), "J/d\n")
cat("Growth:        ", round(fl$p_G, 3), "J/d\n")
cat("Struct. length:", round(fl$L, 3), "cm\n")
cat("Scaled reserve:", round(fl$e, 3), "\n")
```


# Summary

| Analysis | Model type | Key output |
|----------|-----------|------------|
| Single worm growth | `"individual"` | DEB parameter posteriors, PPC |
| 21 worms (population) | `"hierarchical"` | Population mean/SD of $p_{Am}$, shrinkage |
| Toxicant effect | `"debtox"` | EC50, NEC with uncertainty |

The Bayesian framework gives you:

1. **Full uncertainty quantification** — every derived quantity comes
   with a posterior distribution.
2. **Prior incorporation** — biological knowledge informs estimation.
3. **Hierarchical structure** — partial pooling borrows strength across
   individuals.
4. **Model checking** — PPC and diagnostics catch model mis-specification.

## Session info

```{r session}
sessionInfo()
```
